Equalizer Programming Guide

### 2008-07-21 Developer/ProgrammingGuide/paper.tex
Implement automatic image compression on 'slow' links, add accum and samples hint to window attributes. Amend the documentation accordingly.
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[x] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\date{
  \vspace{2cm}
  \begin{figure}[ht]
    \hspace{-4.5cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{figure}
  \vspace{1cm}
%  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
  Version 1.3.5 for Equalizer 0.5.3+svn\\[\medskipamount]
  \today
}

\newcommand{\tm}{\texttrademark~}
\newcommand{\rc}{\raise 1ex\hbox{{\tiny\textregistered}}~}
\newcommand{\fig}[1]{Figure~\ref{#1}}
\newcommand{\sref}[1]{Section~\ref{#1}}
\newcommand{\link}[1]{\htmladdnormallink{#1}{#1}}

@@

The images on the front page show the following Equalizer applications:
a terrain rendering application on a six-node display wall [top left], a
flow visualization application for climate research\footnote{Image
  courtesy of Computer Graphics and Multimedia Systems, University of
  Siegen}, the \textsf{eVolve} volume renderer\footnote{Data sets
  courtesy of General Electric, USA and AVS, USA} [bottom left] the
\textsf{eqPly} polygonal renderer in a three-sided CAVE [bottom center],
and a six-node sort-last database decomposition with parallel
direct-send recomposition\footnote{Data set courtesy of Stanford
  University Computer Graphics Laboratory} [bottom right].

\clearpage
\thispagestyle{empty}
\tableofcontents
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Version for Equalizer 0.5\\
    1.3.1   & May 02, 2008 & Add statistics, improved images based on 1.2 print\\
    1.3.2   & May 05, 2008 & Add network distribution on eqPly model data\\
    1.3.3   & May 10-19, 2008 & File Format Appendix\\
    1.3.4   & July 3, 2008 & Update statistics section\\
    1.3.4   & July 7, 2008 & Update after namespace naming cleanup\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \vspace{-1em}
  \includegraphics[width=.618\textwidth]{images/executionFlow.pdf}
  {\caption{\small\label{fExecutionFlow}Parallel Rendering}}
  \vspace{-1em}
\end{wrapfigure}

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \subfigure[]{\label{fSync}
    \includegraphics[width=.25\textwidth]{images/sync}}\hfil
  \subfigure[]{\label{fAsync}
    \includegraphics[width=.25\textwidth]{images/async}}%
  {\caption{\small\label{fSyncAsync}Synchronous and asynchronous execution}}
\end{wrapfigure}
By default, \textsf{Config::finish\-Fra\-me} also synchronizes the
completion of all local rendering tasks for the current frame. This
facilitates porting of exising rendering codes, since the database does
not have to be multi-buffered. Applications such as \textsf{eqPly}, which
do not need this per-node frame synchronization can disable it, as
explained in \sref{sNodeFrame}.

\fig{fSyncAsync} shows the execution of (hypothetical) rendering tasks
without latency (\fig{fSync}) and with a latency of
one frame (\fig{fAsync}). With \textsf{eqPly}, a speedup of 15\% has been
observed on a five-node rendering cluster when using a latency of one frame
instead of no
latency\footnote{\link{http://www.equalizergraphics.com/scalability.html}}. A
latency of one or two frames is normally not perceived by the user.

@@

Applications distributing a dynamic scene graph use the frame data
instead of the init data as the entry point to their scene graph data
structure. \fig{fObjects} shows one possible implementation, where the
identifier and version of the scene graph root are transported using the
frame data. The scene graph root then serializes and de-serializes his
immediate children by transferring their identifier and current version,
similar to the static distribution done by \textsf{eqPly}.

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/objects.pdf}
  {\caption{\small\label{fObjects}Scene Graph Distribution}}
  \vspace{-4em}
\end{wrapfigure}
The objects are still created by the application, and then registered or
mapped with the session in order to distribute them. When mapping
objects in a hierarchical data structure, their type often has to be
known in order to create them. Equalizer does not currently provide
object typing, this has to be done by the application, either implicitly
in the current implementation context, or by transferring a type
identifier. In \textsf{eqPly}, object typing is implicit since it is
well-defined which object is mapped in which context.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/configInit.pdf}
  {\caption{\small\label{fConfigInit}Config Initialization Sequence}}
\end{wrapfigure}
The tasks are executed by the node and pipe threads in parallel. The
parent's initialization methods are always executed before any child
initialization method. This parallelization is necessary to allow a
speedy startup of the configuration on large-scale graphics clusters. On
the other hand, it means that initialization functions are called even
if the parent's initialization has failed.

@@

If the configuration was initialized correctly, the configuration tries
to set up a tracking device for head tracking. Equalizer does not
provide extensive support for tracking devices, as this is an orthogonal
problem to parallel rendering. Tracking device support has already been
solved by a number of implementations\footnote{VRCO Trackd VRPN, etc.},
which can easily be integrated with Equalizer. The example code in
\textsf{eqPly} provides a reference implementation for the integration
of such a tracking library. \sref{sTracking} provides more background on
head tracking.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/windowInit.pdf}
  {\caption{\small\label{fWindowInit}Window Initialization}}
%  \vspace{-1em}
\end{wrapfigure}
The state object is used to handle the creation of OpenGL objects in a
multipipe, multi-threaded execution environment. It uses an object
manager, which is described in detail in \sref{sObjectManager}. It is
used in conjunction with a reference-counting smart-pointer here, since it is
potentially `owned' by multiple windows at the same time.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \vspace{-1em}
  \includegraphics[width=.618\textwidth]{images/render.pdf}
  {\caption{\small\label{fRender}Main Render Loop}}
  \vspace{-2em}
\end{wrapfigure}
The model data is spatially organized in an 3-dimensional
kD-tree\footnote{See also \link{http://en.wikipedia.org/wiki/Kd-tree}}
for efficient view frustum culling. When the model is loaded by
\textsf{Node::config\-Init}, it is preprocessed into the kD-tree and each
node of the tree gets a database range assigned. The root node
has the range [0, 1], its left child [0, 0.5] and its right child [0.5,
1], and so on for all nodes in the tree. The preprocessed model is saved
in a binary format for accelerating subsequent use.

The rendering loop maintains a list of candidates to render, which
initially contains the root node. Each candidate of this list is tested
for full visiblity against the frustum and range, and rendered if
visible. It is dropped if it is fully invisible or fully out of
range. If it is partially visible or partially in range, the children of
the node are added to the candidate list. \fig{fRender} shows a flow
chart of the rendering algorithm, which performs efficient view frustum
and range culling.

@@

Events are received by an event handler. The event handler finds the
\textsf{eq::Window} associated to the event. It then creates a generic
\textsf{WindowEvent}, which holds important event data in an independent
format. The original event is attached to the generic window event.

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
  {\caption{\small\label{fEventProcessing}Event Processing}}
\end{wrapfigure}
The event handler then passes the window event to
\textsf{Window::processEvent}, which is responsible for either handling
the event locally, or for translating it into a generic
\textsf{ConfigEvent}. The config events are send to the application
thread using \textsf{Config::sendEvent}.

@@

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
  {\caption{\small\label{fDirectSend}Parallel Direct Send Compositing}}
\end{wrapfigure}
The main idea behind direct send is to parallelize the costly
recomposition for database (sort-last) decomposition. With each
additional source channel, the amount of pixel data to be composited
grows linearly. When using the simple approach of compositing all frames
on the destination channel, this channel quickly becomes the bottleneck
in the system. Direct send distributes this workload evenly across all
source channels, and thereby keeps the compositing work per channel
constant.

In direct send compositing, each rendering channel is also responsible
for the sort-last composition of one screen-space tile. He receives
the framebuffer pixels for his tile from all the other channels. The
size of one tile decreases linearly with the number of source channels,
which keeps the total amount of pixel data per channel constant.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/slices.pdf}
  {\caption{\small\label{fSlices}Blending Slices in 3D-Texture-based
      Volume Rendering}}
\end{wrapfigure}
The \textsf{eVolve} example is a scalable volume renderer. It uses 3D
texture-based volume rendering, where the 3D texture is intersected by
view-aligned slices. The slices are rendered back-to-front and blended
together to produce the final image, as shown in
\fig{fSlices}\footnote{Volume Data Set courtesy of: SFB-382 of the German
  Research Council (DFG)}.

When using 2D (sort-first) or stereo decompositions, no special
programming is needed to achieve good scalability, as \textsf{eVolve} is
mostly fill-limited and therefore scales nicely in these modes.

@@

\subsection{Statistics}

@@

\newpage
\appendix
\section{File Format}

@@

@@

The hostname is the IP address or (resolvable) host name. When used for
the server, the server binds its listening socket only to the specified
name.
\clearpage

@@

    attributes
    {
        hint_stereo         OFF | ON | AUTO
        hint_doublebuffer   OFF | ON | AUTO
        hint_decoration     OFF | ON
        hint_fullscreen     OFF | ON
        hint_swapsync       OFF | ON         // AGL, WGL only
        hint_drawable       window | pbuffer
        hint_statistics     off | fastest [on] | nicest
        planes_color        unsigned
        planes_alpha        unsigned
        planes_depth        unsigned
        planes_stencil      unsigned
    }

>>>>

\date{
  \vspace{2cm}
  \begin{figure}[ht]
    \hspace{-4.5cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{figure}
  \vspace{1cm}
%  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
  Version 1.3.6 for Equalizer 0.5.3+svn\\[\medskipamount]
  \today
}

\newcommand{\tm}{\texttrademark~}
\newcommand{\rc}{\raise 1ex\hbox{{\tiny\textregistered}}~}
\newcommand{\fig}[1]{Figure~\ref{#1}}
\newcommand{\sref}[1]{Section~\ref{#1}}
\newcommand{\aref}[1]{Appendix~\ref{#1}}
\newcommand{\link}[1]{\htmladdnormallink{#1}{#1}}

@@

The images on the front page show the following Equalizer applications:
a terrain rendering application on a six-node display wall [top left], a
flow visualization application for climate research\footnote{Image
  courtesy of Computer Graphics and Multimedia Systems, University of
  Siegen}, the \textsf{eVolve} volume renderer\footnote{Data set
  courtesy of General Electric, USA} [bottom left] the \textsf{eqPly}
polygonal renderer in a three-sided CAVE [bottom center], and a six-node
sort-last database decomposition with parallel direct-send
recomposition\footnote{Data set courtesy of Stanford University Computer
  Graphics Laboratory} [bottom right].

\clearpage
\thispagestyle{empty}
\tableofcontents
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Version for Equalizer 0.5\\
    1.3.1   & May 02, 2008 & Add \sref{sStatistics}, improved images based on 1.2 print\\
    1.3.2   & May 05, 2008 & Add network distribution on eqPly model data\\
    1.3.3   & May 10-19, 2008 & add \aref{aFileFormat}\\
    1.3.4   & July 3, 2008 & Update \sref{sStatistics}\\
    1.3.5   & July 7, 2008 & Update after namespace naming cleanup\\
    1.3.6   & July 20, 2008 & Add accum and samples to \aref{aFileFormat}, add \sref{sDrawableConfig}\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/executionFlow.pdf}
  {\caption{\small\label{fExecutionFlow}Parallel Rendering}}
\end{wrapfigure}

@@

By default, \textsf{Config::finish\-Fra\-me} also synchronizes the
completion of all local rendering tasks for the current frame. This
facilitates porting of exising rendering codes, since the database does
not have to be multi-buffered. Applications such as \textsf{eqPly}, which
do not need this per-node frame synchronization can disable it, as
explained in \sref{sNodeFrame}.

\begin{figure}[ht!]\center
  \includegraphics[width=\textwidth]{images/syncAsync}
  {\caption{\small\label{fSyncAsync}Synchronous and asynchronous execution}}
\end{figure}
\fig{fSyncAsync} shows the execution of the rendering tasks of a 2-node
2D compound without latency and with a latency of one frame. The
asynchronous execution pipelines certain rendering operations and hides
imbalances in the load distribution, resulting in an improved
framerate. With \textsf{eqPly}, a speedup of 15\% has been observed on a
five-node rendering cluster when using a latency of one frame instead of
no
latency\footnote{\link{http://www.equalizergraphics.com/scalability.html}}. A
latency of one or two frames is normally not perceived by the user.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/objects.pdf}
  {\caption{\small\label{fObjects}Scene Graph Distribution}}
\end{wrapfigure}
Applications distributing a dynamic scene graph use the frame data
instead of the init data as the entry point to their scene graph data
structure. \fig{fObjects} shows one possible implementation, where the
identifier and version of the scene graph root are transported using the
frame data. The scene graph root then serializes and de-serializes his
immediate children by transferring their identifier and current version,
similar to the static distribution done by \textsf{eqPly}.

The objects are still created by the application, and then registered or
mapped with the session in order to distribute them. When mapping
objects in a hierarchical data structure, their type often has to be
known in order to create them. Equalizer does not currently provide
object typing, this has to be done by the application, either implicitly
in the current implementation context, or by transferring a type
identifier. In \textsf{eqPly}, object typing is implicit since it is
well-defined which object is mapped in which context.

@@

The tasks are executed by the node and pipe threads in parallel. The
parent's initialization methods are always executed before any child
initialization method. This parallelization is necessary to allow a
speedy startup of the configuration on large-scale graphics clusters. On
the other hand, it means that initialization functions are called even
if the parent's initialization has failed.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/configInit.pdf}
  {\caption{\small\label{fConfigInit}Config Initialization Sequence}}
\end{wrapfigure}
If the configuration was initialized correctly, the configuration tries
to set up a tracking device for head tracking. Equalizer does not
provide extensive support for tracking devices, as this is an orthogonal
problem to parallel rendering. Tracking device support has already been
solved by a number of implementations\footnote{VRCO Trackd VRPN, etc.},
which can easily be integrated with Equalizer. The example code in
\textsf{eqPly} provides a reference implementation for the integration
of such a tracking library. \sref{sTracking} provides more background on
head tracking.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/windowInit.pdf}
  {\caption{\small\label{fWindowInit}Window Initialization}}
\end{wrapfigure}
The state object is used to handle the creation of OpenGL objects in a
multipipe, multi-threaded execution environment. It uses an object
manager, which is described in detail in \sref{sObjectManager}. It is
used in conjunction with a reference-counting smart-pointer here, since it is
potentially `owned' by multiple windows at the same time.

@@

The model data is spatially organized in an 3-dimensional
kD-tree\footnote{See also \link{http://en.wikipedia.org/wiki/Kd-tree}}
for efficient view frustum culling. When the model is loaded by
\textsf{Node::config\-Init}, it is preprocessed into the kD-tree and each
node of the tree gets a database range assigned. The root node
has the range [0, 1], its left child [0, 0.5] and its right child [0.5,
1], and so on for all nodes in the tree. The preprocessed model is saved
in a binary format for accelerating subsequent use.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/render.pdf}
  {\caption{\small\label{fRender}Main Render Loop}}
\end{wrapfigure}
The rendering loop maintains a list of candidates to render, which
initially contains the root node. Each candidate of this list is tested
for full visiblity against the frustum and range, and rendered if
visible. It is dropped if it is fully invisible or fully out of
range. If it is partially visible or partially in range, the children of
the node are added to the candidate list. \fig{fRender} shows a flow
chart of the rendering algorithm, which performs efficient view frustum
and range culling.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
  {\caption{\small\label{fEventProcessing}Event Processing}}
  \vspace{-3ex}
\end{wrapfigure}
Events are received by an event handler. The event handler finds the
\textsf{eq::Window} associated to the event. It then creates a generic
\textsf{WindowEvent}, which holds important event data in an independent
format. The original event is attached to the generic window event.

The event handler then passes the window event to
\textsf{Window::processEvent}, which is responsible for either handling
the event locally, or for translating it into a generic
\textsf{ConfigEvent}. The config events are send to the application
thread using \textsf{Config::sendEvent}.

@@

\subsubsection{\label{sDrawableConfig}Drawable Configuration}

OpenGL drawables have a multitude of buffer modes. A drawable might be
single-buffered, double-buffered or quad-buffered, have auxilary image
planes such as stencil, accumulation and depth buffer or multisampling.

The OpenGL drawable is configured using window attributes. These
attributes are used by the method choosing the pixel format (or visual
in X11 speak) to select the correct drawable configuration.

Window attributes can either be configured through the configuration
file (see \aref{aFileFormat}), or programmatically. In the configuration
file, modes are selected which are not application-specific, for example
stereo formats for active stereo displays.

Applications which require certain drawable attributes can set the
corresponding window attribute hint during window initialization. The
Equalizer volume rendering example, \textsf{eVolve}, is such an
example. It does need alpha planes for rendering and compositing. The
window initialization of \textsf{eVolve} sets the attribute before
calling the default initialization method of Equalizer:

{\footnotesize\begin{lstlisting}
bool eVolve::Window::configInit( const uint32_t initID )
{
    // Enforce alpha channel, since we need one for rendering
    setIAttribute( IATTR_PLANES_ALPHA, 8 );

    return eq::Window::configInit( initID );
}
\end{lstlisting}}

@@

The main idea behind direct send is to parallelize the costly
recomposition for database (sort-last) decomposition. With each
additional source channel, the amount of pixel data to be composited
grows linearly. When using the simple approach of compositing all frames
on the destination channel, this channel quickly becomes the bottleneck
in the system. Direct send distributes this workload evenly across all
source channels, and thereby keeps the compositing work per channel
constant.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
  {\caption{\small\label{fDirectSend}Parallel Direct Send Compositing}}
\end{wrapfigure}
In direct send compositing, each rendering channel is also responsible
for the sort-last composition of one screen-space tile. He receives
the framebuffer pixels for his tile from all the other channels. The
size of one tile decreases linearly with the number of source channels,
which keeps the total amount of pixel data per channel constant.

@@

The \textsf{eVolve} example is a scalable volume renderer. It uses 3D
texture-based volume rendering, where the 3D texture is intersected by
view-aligned slices. The slices are rendered back-to-front and blended
together to produce the final image, as shown in
\fig{fSlices}\footnote{Volume Data Set courtesy of: SFB-382 of the German
  Research Council (DFG)}.

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/slices.pdf}
  {\caption{\small\label{fSlices}Blending Slices in 3D-Texture-based
      Volume Rendering}}
\end{wrapfigure}
 When using 2D (sort-first) or stereo decompositions, no special
programming is needed to achieve good scalability, as \textsf{eVolve} is
mostly fill-limited and therefore scales nicely in these modes.

@@

\subsection{\label{sStatistics}Statistics}

@@

\newpage
\appendix
\section{\label{aFileFormat}File Format}

@@

\begin{tabularx}{\textwidth}{|l|X|}
  \hline
  \textbf{Name} & EQ\_WINDOW\_IATTR\_PLANES\_ACCUM\\
  \textbf{Value} & unsigned\\
  \textbf{Default} & UNDEFINED\\
  \textbf{Details} & Determines the number of color accumulation buffer
  planes for the window. The interpretation of this value is window
  system-specific, as some window systems select a visual with the
  closest match to this value, and some select a visual with at least
  the number of accumulation buffer planes specified.\\
  \hline
\end{tabularx}\\\vfill

\begin{tabularx}{\textwidth}{|l|X|}
  \hline
  \textbf{Name} & EQ\_WINDOW\_IATTR\_PLANES\_ACCUM\_ALPHA\\
  \textbf{Value} & unsigned\\
  \textbf{Default} & UNDEFINED\\
  \textbf{Details} & Determines the number of alpha accumulation buffer
  planes for the window. The interpretation of this value is window
  system-specific, as some window systems select a visual with the
  closest match to this value, and some select a visual with at least
  the number of accumulation buffer planes specified. If this attribute
  is undefined, the value of EQ\_WINDOW\_IATTR\_PLANES\_ACCUM is used to
  determine the number of alpha accumulation buffer planes.\\
  \hline
\end{tabularx}\\\vfill

\begin{tabularx}{\textwidth}{|l|X|}
  \hline
  \textbf{Name} & EQ\_WINDOW\_IATTR\_PLANES\_SAMPLES\\
  \textbf{Value} & unsigned\\
  \textbf{Default} & UNDEFINED\\
  \textbf{Details} & Determines the number of samples used for multisampling.\\
  \hline
\end{tabularx}\\\vfill

@@

The hostname is the IP address or (resolvable) host name. When used for
the server, the server binds its listening socket only to the specified
name.

@@

    attributes
    {
        hint_stereo         OFF | ON | AUTO
        hint_doublebuffer   OFF | ON | AUTO
        hint_decoration     OFF | ON
        hint_fullscreen     OFF | ON
        hint_swapsync       OFF | ON         // AGL, WGL only
        hint_drawable       window | pbuffer
        hint_statistics     off | fastest [on] | nicest
        planes_color        unsigned
        planes_alpha        unsigned
        planes_depth        unsigned
        planes_stencil      unsigned
        planes_accum        unsigned
        planes_accum_alpha  unsigned
        planes_samples      unsigned
    }
